---
title: "COVID-19 Data Analysis"
author: "Peter Tarara"
date: "2022-10-08"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lubridate)
```

## Data Source

I will be using data from the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University for my analysis. The repository contains data related to COVID cases and can be found here: https://github.com/CSSEGISandData/COVID-19

## Data Summary and Tidying

```{r import, echo=TRUE}
url_in <- 
  paste0("https://raw.githubusercontent.com/",
         "CSSEGISandData/COVID-19/master/",
         "csse_covid_19_data/csse_covid_19_time_series/")

file_names <-
  c("time_series_covid19_confirmed_global.csv",
    "time_series_covid19_deaths_global.csv",
    "time_series_covid19_confirmed_US.csv",
    "time_series_covid19_deaths_US.csv")

urls <- str_c(url_in, file_names)

global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])

global_cases <- global_cases %>%
     pivot_longer(cols = -c(`Province/State`,
                            `Country/Region`, Lat, Long),
                  names_to = "date",
                  values_to = "cases") %>% 
  select(-c(Lat,Long))


global_deaths <- global_deaths %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat, Long))

global <- global_cases %>%
  full_join(global_deaths) %>%
  mutate(date = mdy(date))

global <- global %>% filter(cases > 0) #issue with this filter

summary(global)

us_cases <- us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us_deaths <- us_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US <- us_cases %>%
  full_join(us_deaths)

global <- global %>%
  unite("Combined_Key",
        c(`Province/State`, `Country/Region`),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

uid_lookup_url <- 
  paste0("https://raw.githubusercontent.com/",
  "CSSEGISandData/COVID-19/master/csse_covid_",
  "19_data/UID_ISO_FIPS_LookUp_Table.csv")

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))


global <- global %>% rename(`Country_Region` = `Country/Region`)
global <- global %>% rename(`Province_State` = `Province/State`)

global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```


## Visuals

```{r visuals, echo=TRUE}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths,deaths_per_mill,
         Population) %>%
  ungroup()

US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_state_totals <-US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases), Population = max(Population)) %>%
  filter(cases > 0, Population > 0)


US_state_totals %>%
  mutate(Province_State = fct_reorder(Province_State, cases)) %>%
  filter(cases > 0) %>%
  ggplot(aes(x=Province_State, y=cases)) +
  geom_bar(stat="identity") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 Cases by State")


US_state_totals %>%
  mutate(Province_State = fct_reorder(Province_State, cases)) %>%
  filter(cases > 0) %>%
  ggplot(aes(x=Province_State, y=Population)) +
  geom_bar(stat="identity") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "Population by State")

```

## Anaylsis

When comparing the above two charts, it appears that the population count of a state is useful for predicting the total number of COVID cases. Although this is an obvious observation, I was curious to see if any outliers could raise more questions, such as a clear difference between states with various approaches to virus containment. To test this theory, I used a linear model to visualize the relationship between population and the number of COVID cases.

## Model

```{r model, echo=TRUE}
mod <- lm(cases ~ Population, data = US_state_totals)

US_state_totals %>% mutate(pred = predict(mod))

US_state_totals_w_pred <- US_state_totals %>% mutate(pred = predict(mod))

US_state_totals_w_pred %>% ggplot() +
geom_point(aes(x= Population, y = cases), color = "blue") +
 geom_point(aes(x= Population, y = pred), color = "red")
```

## Conclusion and Bias

In the above chart, the red dots are the predictions, and the blue dots are the actual values. Based on this model, I believe it is fair to conclude that population is a valuable predictor of the total number of COVID cases per state.

My bias for this analysis would be my expectation that a greater population would lead to a higher number of COVID cases.
